---
title: "16S Phyloseq"
output: html_notebook
---
Load required packages.
```{r}
require(phyloseq)
 require(parallelDist)
require(ggplot2)
  require(randomForest)
require(Boruta)
  require(phyloseq)
require(stats)
  require(dplyr)
require(readr)
  require(pheatmap)
require(tidyverse)
  require(micrUBIfuns)
require(circlize)
  require(phylosmith)
```

Load necessary files from Dada2 processing script. 
```{r}
load("/Users/gordoncuster/Desktop/Git_Projects/BeeBread/Data/16S/DADA2outputs/dada2_outputs_16S_BeeBread_wtree.RData")
```

Read in metadata 
```{r}
metadata<-read.csv("/Users/gordoncuster/Desktop/Git_Projects/BeeBread/Data/BB_sample-metadata.txt", sep = "\t")
rownames(metadata)<-metadata$X
metadata$Region_Site<-paste( metadata$Region, metadata$Site, sep = "")
metadata$Treatment_Site<-paste( metadata$Treatment, metadata$Site, sep = "")
```

Convert files to correct format for phyloseq object and then rename samples in OTU table to match the  format of the metadata file. Create phyloseq object and root phylogenetic tree for usage with distance metrics (e.g., Unifrac). 
```{r}
#assign object type to merge into phyloseq object
md<-sample_data(metadata)
otu<-otu_table(seqtab.nochim, taxa_are_rows = F)
tax_tab<-tax_table(taxa)
#check sample names of both otu table and metadata
#remove .fastq.gz from otu table names
sample_names(otu)<-str_split(sample_names(otu), pattern = ".fastq.gz", simplify = T) [,1]
#since they match, you can create the phyloseq object
bb16S_orig<-phyloseq(md, otu, tax_tab, fitGTR$tree)

#root tree for distance metrics like unifrac
set.seed(11)
phy_tree(bb16S_orig)<-root(phy_tree(bb16S_orig), sample(taxa_names(bb16S_orig), 1), resolve.root = TRUE)
is.rooted(phy_tree(bb16S_orig))
```

Pre-processing. Rename OTU IDS to something more manageable. We save the ASCV sequences in case we need them later (e.g., BLASTN searches). Create Bacterial and Archaeal phyloseq objects, but first we remove Chloroplasts. 
```{r}
#Extract original IDs for future reference. 
seq_df_w_OTUID<-data.frame(OTUID = 1:ntaxa(bb16S_orig), Sequence = taxa_names(bb16S_orig))
taxa_names(bb16S_orig)<-paste("OTU_" , 1:ntaxa(bb16S_orig), sep = "")

#remove chloroplast
#60 chloroplast taxa
bb16S_orig_nc<-subset_taxa(bb16S_orig, Class != "Chloroplast")

#2257 bacterial taxa - 9 non-bacteria removed. 
bb16S_bac<-subset_taxa(bb16S_orig_nc, Kingdom == "Bacteria")
#9 archaea
bb16S_arch<-subset_taxa(bb16S_orig_nc, Kingdom == "Archaea")
```

Remove control samples as there is only as single replicate of each. 
```{r}
ps_wo_control<-subset_samples(bb16S_bac, Treatment != "none")
```

Examine rarefaction curves. Plateau in most samples under the 5k read mark. 
```{r}
rarecurve(otu_table(ps_wo_control), step=50, cex=0.5)
```

First, check sample depths for rarefaction. Minimum sample depth is a bit over 21k reads, so we will rarefy at 20k.  
```{r}
set.seed(11)
sort(sample_sums(ps_wo_control))
bb16S_bac_rarefy<-rarefy_even_depth(ps_wo_control, sample.size = 20000, rngseed = 14, trimOTUs = T)
#300 otus were removed during rarefaction. 

#To move forward, we can hellinger transform and go from there.
bb16S_bac_hellinger<-transform_sample_counts(ps_wo_control, function(x) sqrt(x / sum(x)))

#maximum liklihood point estimates
bb16S_bac_ML<- transform_sample_counts(ps_wo_control, function(x) x / sum(x))
```

#Exploratory analysis of Alpha diversity

Plots of Alpha Diversity by region, treatment, timepoint, and the combinations. 
```{r}
plot_richness(bb16S_bac_rarefy, x = "Region", measures = c("Shannon", "Observed", "Chao1"), color = "Region") + geom_boxplot() + ggtitle("Region")

plot_richness(bb16S_bac_rarefy, x = "Treatment", measures = c("Shannon", "Observed", "Chao1"), color = "Treatment") + geom_boxplot() + ggtitle("Treatment")

plot_richness(bb16S_bac_rarefy, x = "Timepoint", measures = c("Shannon", "Observed", "Chao1"), color = "Timepoint") + geom_boxplot() + ggtitle("Timepoint")

plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Shannon", "Observed", "Chao1"), color = "Site") + geom_boxplot() + ggtitle("Site")

plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Shannon", "Observed", "Chao1"), color = "Timepoint") + geom_boxplot() + ggtitle("Site by timepoint")

plot_richness(bb16S_bac_rarefy, x = "Region_Treatment", measures = c("Shannon", "Observed", "Chao1"), color = "Region_Treatment") + geom_boxplot() + ggtitle("Region_Treatment")

plot_richness(bb16S_bac_rarefy, x = "Region_Site", measures = c("Shannon", "Observed", "Chao1"), color = "Region_Site") + geom_boxplot() + ggtitle("Region_site")

plot_richness(bb16S_bac_rarefy, x = "Region_Timepoint", measures = c("Shannon", "Observed", "Chao1"), color = "Region_Timepoint") + geom_boxplot() + ggtitle("Region_Timepoint")

plot_richness(bb16S_bac_rarefy, x = "Treatment_Timepoint", measures = c("Shannon", "Observed", "Chao1"), color = "Treatment_Timepoint") + geom_boxplot() + ggtitle("Treatment_Timepoint")
```
Based upon the site and region graphs, it appears that site contains more differences than region. There are no obvious differences among treatments. The site by timepoint graph shows what appears to be an interaction of before and after treatment and site. Some sites display lower diversity following treatment and others higher, with consistent trends across all sites though. 


Statistical testing of alpha diversity metrics. Using ANOVA framework and checking for assumption of normality and correlation in predictors. region and site appear to be associated, so I will drop Region for this exploratory analysis. We can revisit if we decide Region is a more accurate predictor, but it appears that from the graphs above site provides more resolution than region.
https://www.pluralsight.com/guides/testing-for-relationships-between-categorical-variables-using-the-chi-square-test
```{r}
#create alpha diversity table and prep data to include grouping columns
richness_BB16S_bac<-estimate_richness(bb16S_bac_rarefy, measures = c("Shannon", "Observed", "Chao1"))
#add metadata
richness_BB16S_bac$Region<-sample_data(bb16S_bac_rarefy)$Region
richness_BB16S_bac$Site<-sample_data(bb16S_bac_rarefy)$Site
richness_BB16S_bac$Timepoint<-sample_data(bb16S_bac_rarefy)$Timepoint
richness_BB16S_bac$Treatment<-sample_data(bb16S_bac_rarefy)$Treatment
#add combo columns
richness_BB16S_bac$Region_Treatment<-sample_data(bb16S_bac_rarefy)$Region_Treatment
richness_BB16S_bac$Region_Site<-sample_data(bb16S_bac_rarefy)$Region_Site
richness_BB16S_bac$Region_Timepoint<-sample_data(bb16S_bac_rarefy)$Region_Timepoint
richness_BB16S_bac$Treatment_Timepoint<-sample_data(bb16S_bac_rarefy)$Treatment_Timepoint

#test of association between 
chisq.test(richness_BB16S_bac$Region, richness_BB16S_bac$Site, simulate.p.value = TRUE)
```

Tests of Shannon diversity - significant differences across sampling sites and interaction term of timepoint and site. 
```{r}
#ANOVA and Tukey's PW comparisons
#Run a full 3 way ANOVA with interactions
mod<-aov(Shannon ~ Site * Treatment * Timepoint, data = richness_BB16S_bac)
#residuals look fine. 
shapiro.test(mod$residuals)
summary(mod)
#Summary of model shows site to be only lower order significant predictor.
#interaction term of site x time point as well. - this was picked up in the exploratory graphs. 

#What about pairwise differences in site
TukeyHSD(mod, "Site")
plot(TukeyHSD(mod, "Site"), las=1, cex.axis = 0.4)

#Several significant pairwise differences to accompany the significant global test for interaction term. We can
TukeyHSD(mod, "Site:Timepoint")
plot(TukeyHSD(mod, "Site:Timepoint"), las=1, cex.axis = 0.4)

plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Shannon"), color = "Timepoint") + geom_boxplot() + ggtitle("Site x Timepoint")
```

Tests of Chao 1 - emphasizes importance of rare taxa. Site is significant as are the interactions with site. 
```{r}
#ANOVA and Tukey's PW comparisons
#Run a full 3 way ANOVA with interactions
mod<-aov(Chao1 ~ Site * Treatment * Timepoint, data = richness_BB16S_bac)
#residuals look fine. 
shapiro.test(mod$residuals)
summary(mod)
#Site and the interactions with site are significant. 

TukeyHSD(mod, "Site")
plot(TukeyHSD(mod, "Site"), las=1, cex.axis = 0.4)

plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Chao1"), color = "Timepoint") + geom_boxplot() + ggtitle("Site x Timepoint")
plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Chao1"), color = "Treatment") + geom_boxplot() + ggtitle("Site x Treatment")
```

Tests of Observed - unique taxa in a sample. Site is significant as are the interactions with site. 
```{r}
#ANOVA and Tukey's PW comparisons
#Run a full 3 way ANOVA with interactions
mod<-aov(Observed ~ Site * Treatment * Timepoint, data = richness_BB16S_bac)
#residuals look fine. 
shapiro.test(mod$residuals)
summary(mod)
#Summary of model shows site to be a significant predictor as well as the site interaction terms.
TukeyHSD(mod, "Site")
plot(TukeyHSD(mod, "Site"), las=1, cex.axis = 0.4)

plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Observed"), color = "Timepoint") + geom_boxplot() + ggtitle("Site")
plot_richness(bb16S_bac_rarefy, x = "Site", measures = c("Observed"), color = "Treatment") + geom_boxplot() + ggtitle("Site")
```
Summary of alpha diversity: Rarefying at 20k reads sufficiently captures the diversity of each sample, as demonstrated by the plateau in our rarefaction curves. Site is the most useful predictor for determining differences in alpha diversity. Several interactions with site are present for all metrics of alpha diversity. No clear trends in alpha diversity differences. Some sites display shifts responding to treatment, while others do not. Some sites show increased alpha diversity at timepoint 2 and others the opposite. 



#Multivaraite exploratory analysis

Info on strata argument. Briefly, this allows us to restrict permutations to within a group and control for an effect in an analogous way to a random effect. 

https://stats.stackexchange.com/questions/350462/can-you-perform-a-permanova-analysis-on-nested-data/350504#350504?newreg=8e389dc237c54d87ab240486b8b0fe33
https://stats.stackexchange.com/questions/188519/adonis-in-vegan-order-of-variables-or-use-of-strata?noredirect=1&lq=1
https://stats.stackexchange.com/questions/459407/permanova-outputs-with-or-without-random-factor
https://ichthyology.usm.edu/courses/multivariate/feb_7.pdf 

Both with Bray-curtis and Unifrac distances. 
```{r}
#extract data from phyloseq object
sd_adonis<-data.frame(sample_data(bb16S_bac_ML))
sd_adonis$Treatment <-as.factor(sd_adonis$Treatment)
sd_adonis$Site <-as.factor(sd_adonis$Site)
dist<-phyloseq::distance(bb16S_bac_ML, method =  "bray")

#full model with no strata arguemnt. Essentiall the effect of each main effect without controlling for any potential groups. 
adonis(dist ~  Site * Treatment * Timepoint, data= sd_adonis, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations to time points. 
adonis(dist ~  Site * Treatment * Timepoint, data= sd_adonis, strata = sd_adonis$Timepoint, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations sites. 
adonis(dist ~  Site * Treatment * Timepoint, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)

#another way to specify the same model using adonis2. Both provide same results as would be expected. No need to run again, merely a sanity check.
#perm<-how(nperm = 10000)
#setBlocks(perm)<-with(sd_adonis, Timepoint)
#adonis2(dist ~  Site * Treatment * Timepoint, data= sd_adonis, permutations = perm, by = "terms")

#weighted unifrac distance
dist.uf<-phyloseq::distance(bb16S_bac_ML, method =  "wunifrac")
#full model with no strata argument, essentially the effect of each main effect without controlling for any potential groups. 
adonis(dist.uf ~  Site * Treatment * Timepoint, data= sd_adonis, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations to time points. 
adonis(dist.uf ~  Site * Treatment * Timepoint, data= sd_adonis, strata = sd_adonis$Timepoint, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations sites. 
adonis(dist.uf ~  Site * Treatment * Timepoint, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)

```
Based on results, site is the main determinant of microbiome composition (r2 of 0.51 and r2 of 0.38, BC and weighted unifrac, respectively). Time point is also significant but explains much less of the variation (r2 of 0.063 and r2 of 0.021, BC and weighted unifrac, respectively). Treatment is only significant for BC (r2 of 0.024). Both remain significant for BC dissimilarity when site and timepoint are controlled for using the strata argument. However, when site is controled for using weighted unifrac dissimilarity, the only significant predictor is the site x treatment interaction. Next, I will split by time point and examine the effect of each treatment while controlling for site effects using the strata argument (strata = site). 

Take away: Site is the strongest predictor of community dissimilarity. This is true for both Bray-Curtis and weighted unifrac distances. 

Split by time point as these are before and after treatment.
```{r}
#split then extract info from each
#pre-treatment samples
bb16S_bac_ML_A<-subset_samples(bb16S_bac_ML, Timepoint == "A")

#extract data from phyloseq object for pre-treatment
sd_adonis<-data.frame(sample_data(bb16S_bac_ML_A))
sd_adonis$Treatment <-as.factor(sd_adonis$Treatment)
sd_adonis$Site <-as.factor(sd_adonis$Site)
dist<-phyloseq::distance(bb16S_bac_ML_A, method =  "bray")

#full models with no strata 
adonis(dist ~  Site * Treatment, data= sd_adonis, permutations = 10000)
adonis(dist ~   Treatment * Site, data= sd_adonis, permutations = 10000)
#models with strata to control for site differences
adonis(dist ~  Site * Treatment, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)
adonis(dist ~  Treatment * Site, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)

#weighted unifrac distance
dist.uf<-phyloseq::distance(bb16S_bac_ML_A, method =  "wunifrac")
#full model with no strata argument, essentially the effect of each main effect without controlling for any potential groups. 
adonis(dist.uf ~  Site * Treatment,  data= sd_adonis, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations sites. 
adonis(dist.uf ~  Site * Treatment, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)
```
Time point 1 (pre-treatment): Significant effect of Site (r2 of 0.73). Treatment is also a significant predictor (r2 of 0.04). This is something to note because this is supposed to be before treatment was applied. When permutations are restricted to within site, treatment and site remain significant. For weighted unifrac, treatment is insignificant (r2 of 0.025) whether site is controlled for or not. 


```{r}
#post-treatment samples
bb16S_bac_ML_B<-subset_samples(bb16S_bac_ML, Timepoint == "B")

#extract data from phyloseq object for post-treatment
sd_adonis<-data.frame(sample_data(bb16S_bac_ML_B))
sd_adonis$Treatment <-as.factor(sd_adonis$Treatment)
sd_adonis$Site <-as.factor(sd_adonis$Site)
dist<-phyloseq::distance(bb16S_bac_ML_B, method =  "bray")

#full models with no strata 
adonis(dist ~  Site * Treatment, data= sd_adonis, permutations = 10000)
adonis(dist ~   Treatment * Site, data= sd_adonis, permutations = 10000)
#models with strata to control for site differences
adonis(dist ~  Site * Treatment, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)
adonis(dist ~  Treatment * Site, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)

#weighted unifrac distance
dist.uf<-phyloseq::distance(bb16S_bac_ML_A, method =  "wunifrac")
#full model with no strata argument, essentially the effect of each main effect without controlling for any potential groups. 
adonis(dist.uf ~  Site * Treatment,  data= sd_adonis, permutations = 10000)
#this model explores the effect of Site, treatment and time point and restricts permutations sites. 
adonis(dist.uf ~  Site * Treatment, data= sd_adonis, strata = sd_adonis$Site, permutations = 10000)
```
Time point 2 (post-treatment): Site was again a significant predictor (r2 of 0.60) and treatment was borderline significant (r2 of 0.03). However, when the effect of site was controlled for by using strata = site, treatment becomes significant (r2 of 0.039). So when controlling for site differences, the treatments contribute to differences in microbiome structure. This was not the case for weighted unifrac dissimilarity. Treatment was not a significant predictor of community dissimilarity regardless of whether or not site differences were controlled for. 

Take away: Site remains the most useful predictor of community dissimilarity. Treatment effects are only observed for non-phylogentic metrics of dissimilarity when the effect of site is controlled for. 


#Visualize ordinations using two methods. CAP and NMDS
```{r}
#extract data from phyloseq object and run anova.cca and drop1(cca) using vegan to examine significance of constraining variables
sd_adonis<-data.frame(sample_data(bb16S_bac_ML))
sd_adonis$Treatment <-as.factor(sd_adonis$Treatment)
sd_adonis$Site <-as.factor(sd_adonis$Site)
tab_adonis<-data.frame(otu_table(bb16S_bac_ML))
#run model and check global significance using anova.cca.
cca_mod<-cca(tab_adonis ~ Site + Timepoint + Treatment, sd_adonis)
anova.cca(cca_mod)
#significant global model so lets look at individual parameters
drop1(cca_mod, test = "perm", permutations = 100)
0#Site and time point are significant but treatment is not. We will not include treatment in our constrained ordination.
```

Constrained ordination using CAP
```{r}
#Bray-Curtis
ord_cap<-ordinate(physeq = bb16S_bac_ML, method = "CAP", distance = "bray", formula = ~  Site + Timepoint )
plot_ordination(physeq = bb16S_bac_ML, ordination = ord_cap, type = "Samples", color = "Site", shape = "Timepoint" ) + theme_classic() + ggtitle("Bray-Curtis CAP")
#strongest grouping by Site. Time point also has a smaller effect and the groupings are not nearly as clear. 

ord_cap<-ordinate(physeq = bb16S_bac_ML, method = "CAP", distance = "wunifrac", formula = ~  Site + Timepoint )
plot_ordination(physeq = bb16S_bac_ML, ordination = ord_cap, type = "Samples", color = "Site", shape = "Timepoint" ) + theme_classic()  + ggtitle("Weighted Unifrac CAP")
#strongest grouping by Site. Time point also has a smaller effect and the groupings are not nearly as clear. There are several outlier samples that ordinate far to the right on the CAP1 axis. 
```

Unconstrained ordination using NMDS
```{r}
#Bray-Curtis
ord_nmds<-ordinate(physeq = bb16S_bac_ML, method = "NMDS", distance = "bray", k = 3, trymax = 1000)
ord_nmds$stress
plot_ordination(physeq = bb16S_bac_ML, ordination = ord_nmds, type = "Samples", color = "Site", shape = "Timepoint" ) + theme_classic() + ggtitle("Bray-Curtis NMDS")
#strongest grouping by Site. Time point also has a smaller effect, but these groupings are more clear than in the CAP ordination. 

#weighted Unifrac
ord_nmds<-ordinate(physeq = bb16S_bac_ML, method = "NMDS", distance = "wunifrac")
ord_nmds$stress
plot_ordination(physeq = bb16S_bac_ML, ordination = ord_nmds, type = "Samples", color = "Site", shape = "Timepoint" ) + theme_classic() + ggtitle("Weighted Unifrac NMDS")
#strongest grouping by Site. Time point also has a smaller effect, but these groupings are more clear than in the CAP ordination. There are several outlier samples that ordinate far to the right on the NMDS1 axis. 
```

#Indicator taxa using Baruta Algorith
Specify function for Baruta that accepts a phyloseq object. 
```{r}
baruta_phyloseq<-function(physeq = "Phyloseq object", GroupingVar = "Grouping variable"){
  #prep data
  df<-data.frame((otu_table(physeq)))
  md<-data.frame(sample_data(physeq))
  tax<-data.frame(tax_table(physeq))
  rownames(tax)<-str_replace(rownames(tax), pattern = "[[.]]", replacement = "=")
  #run simper
 
                  boruta_layer<-Boruta(df , factor(md[,GroupingVar]), 
                      doTrace = 1, ntree = 1000, maxRuns = 1000) # adjust factor!!!!

                  B1<-data.frame(boruta_layer$finalDecision)  
                  B2<-data.frame(apply(boruta_layer$ImpHistory, 2, mean))
                  names(B1)<-"Layer_decision"
                  names(B2)<-"Layer_importance"
                  BB1<-merge(B1, B2, by="row.names", all.x=TRUE)
                  #sorting the results
                  BB1_sort<-BB1[order(BB1$Layer_importance, decreasing = TRUE),] ## this is to give an overview and choose cutoff
                  #picking up important with cutoff
                  BB1_sort_cutoff<-filter(BB1, Layer_importance >= 2) ## number is your importance cutoff

                        #picking up OTUs from full otutable
                        signOTU<-as.matrix(BB1_sort_cutoff$Row.names) #vector with OTUs
                        signOTU<-str_replace(signOTU, pattern = "[[.]]", replacement = ".")
                        #my_taxa
                        tax$OTU <-rownames(tax)
                        signOTU<-str_replace(signOTU, pattern = "[[.]]", replacement = "=")
                        signmy_taxa<-filter(tax, OTU %in% signOTU) 
                        signmy_taxa<-as.data.frame(signmy_taxa)
                        taxa_keep<-signmy_taxa$OTU
return(taxa_keep)
}
```

Run Baruta algorith with full data set. We are looking for taxa that explain differences among sites. 
```{r}
baruta_Site_indicators<-baruta_phyloseq(physeq = bb16S_bac_ML, GroupingVar = "Site")
baruta_Site_indicators_ps_object<-subset_taxa(bb16S_bac_ML, taxa_names(bb16S_bac_ML) %in% baruta_Site_indicators)
```

https://jokergoo.github.io/ComplexHeatmap-reference/book/
Heatmap to visualize differnces with all samples together. 
```{r}
#create subset phyloseq object and pull out taxonomic info
baruta_Site_indicators_ps_object_tax<-tax_table(baruta_Site_indicators_ps_object)
baruta_Site_indicators_ps_object_tax<-data.frame(baruta_Site_indicators_ps_object_tax)
#rename NA as unknown
baruta_Site_indicators_ps_object_tax$Class[is.na(baruta_Site_indicators_ps_object_tax$Class)]<-"Unknown"
baruta_Site_indicators_ps_object_tax$Family[is.na(baruta_Site_indicators_ps_object_tax$Family)]<-"Unknown"
baruta_Site_indicators_ps_object@tax_table<-tax_table(baruta_Site_indicators_ps_object_tax)
fam_names<-data.frame(gsub("column", "", tax_table(baruta_Site_indicators_ps_object)[,3]))

#create color pallete. This goes from green to white depending on z.score of individual taxa
col_fun<-colorRamp2(c(-2, 0, 2), c("green", "white", "red"))
#Heatmap with taxa names
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = tax_table(baruta_Site_indicators_ps_object)[,4], row_dend_reorder =TRUE, col= col_fun)
#heatmap with otuIDs
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = taxa_names(baruta_Site_indicators_ps_object), row_dend_reorder =TRUE, col= col_fun)
```
Based off of initial figure, it looks like there is a strong grouping based on Baruta ID'ed taxa, and there is pattern for every other row. This indicates differences in time points within each sites. We might want to split by time point and then repeat analysis.  

Since we observe what appears to be a time point signature, we can split by time point and rerun the Baruta algorithm. 
```{r}
#pre-treatment samples
bb16S_bac_ML_A<-subset_samples(bb16S_bac_ML, Timepoint == "A")
#run algo and subset phyloseq object
baruta_Site_indicators_A<-baruta_phyloseq(physeq = bb16S_bac_ML_A, GroupingVar = "Site")
baruta_Site_indicators_ps_object_A<-subset_taxa(bb16S_bac_ML_A, taxa_names(bb16S_bac_ML_A) %in% baruta_Site_indicators_A)
# pull out taxonomic info
baruta_Site_indicators_ps_object_tax_A<-tax_table(baruta_Site_indicators_ps_object_A)
baruta_Site_indicators_ps_object_tax_A<-data.frame(baruta_Site_indicators_ps_object_tax_A)
#rename NA as unknown
baruta_Site_indicators_ps_object_tax_A$Class[is.na(baruta_Site_indicators_ps_object_tax_A$Class)]<-"Unknown"
baruta_Site_indicators_ps_object_tax_A$Family[is.na(baruta_Site_indicators_ps_object_tax_A$Family)]<-"Unknown"
#pull out taxonomy table for naming purposes
baruta_Site_indicators_ps_object_A@tax_table<-tax_table(baruta_Site_indicators_ps_object_tax_A)
fam_names<-data.frame(gsub("column", "", tax_table(baruta_Site_indicators_ps_object_A)[,3]))

#heatmap with taxa names
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_A, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = tax_table(baruta_Site_indicators_ps_object_A)[,4], row_dend_reorder =TRUE, col= col_fun)
#heatmap with otuIDs
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_A, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = taxa_names(baruta_Site_indicators_ps_object_A), row_dend_reorder =TRUE, col= col_fun)
```
Strong grouping of sites based on Baruta ID'ed taxa at the first sampling. 

```{r}
#post-treatment samples
bb16S_bac_ML_B<-subset_samples(bb16S_bac_ML, Timepoint == "B")
#run algo and subset phyloseq object
baruta_Site_indicators_B<-baruta_phyloseq(physeq = bb16S_bac_ML_B, GroupingVar = "Site")
baruta_Site_indicators_ps_object_B<-subset_taxa(bb16S_bac_ML_B, taxa_names(bb16S_bac_ML_B) %in% baruta_Site_indicators_B)
#extract taxonomy info and rename NAs as unknown
baruta_Site_indicators_ps_object_tax_B<-tax_table(baruta_Site_indicators_ps_object_B)
baruta_Site_indicators_ps_object_tax_B<-data.frame(baruta_Site_indicators_ps_object_tax_B)
baruta_Site_indicators_ps_object_tax_B$Class[is.na(baruta_Site_indicators_ps_object_tax_B$Class)]<-"Unknown"
baruta_Site_indicators_ps_object_tax_B$Family[is.na(baruta_Site_indicators_ps_object_tax_B$Family)]<-"Unknown"
baruta_Site_indicators_ps_object_B@tax_table<-tax_table(baruta_Site_indicators_ps_object_tax_B)
fam_names<-data.frame(gsub("column", "", tax_table(baruta_Site_indicators_ps_object_B)[,3]))

#heatmap with taxa names
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = tax_table(baruta_Site_indicators_ps_object_B)[,4], row_dend_reorder =TRUE, col= col_fun)
#heatmap with OTUids
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = taxa_names(baruta_Site_indicators_ps_object_B), row_dend_reorder =TRUE, col= col_fun)
```
Again, strong groupings of sites by Baruta ID'ed taxa at time point 2. 


Next, we will look for taxa that consistently explain the differences among treatments following application. Again we will use the Baruta algorithm, but this time, we will specify the grouping variable as Treatment.

Merge by treatments to see if there are consistencies across treatments. No significant indicators.  
```{r}
#subset to post-treatment samples only
bb16S_bac_ML_B<-subset_samples(bb16S_bac_ML, Timepoint == "B")
#summarize by Treatment_Site
bb16S_bac_ML_B_merged<-merge_samples(bb16S_bac_ML_B, "Treatment")
sample_data(bb16S_bac_ML_B_merged)$Treatment<-rownames(sample_data(bb16S_bac_ML_B_merged))
#run algo and filter phyloseq object
bb16S_bac_ML_B_merged<-prune_taxa(taxa_sums(bb16S_bac_ML_B_merged) > 0, bb16S_bac_ML_B_merged)
baruta_Site_indicators_B_merged<-baruta_phyloseq(physeq = bb16S_bac_ML_B_merged, GroupingVar = "Treatment")
#no taxa indicative of treatment x site interaction.
#Stop this inquiry here. Do not run below because no taxa are identified. 

baruta_Site_indicators_ps_object_B_merged<-subset_taxa(bb16S_bac_ML_B_merged, taxa_names(bb16S_bac_ML_B_merged) %in% baruta_Site_indicators_B_merged)
#extract taxonomy and rename NA to unknown. 
baruta_Site_indicators_ps_object_tax_B_merged<-tax_table(baruta_Site_indicators_ps_object_B_merged)
baruta_Site_indicators_ps_object_tax_B_merged<-data.frame(baruta_Site_indicators_ps_object_tax_B_merged)
baruta_Site_indicators_ps_object_tax_B_merged$Class[is.na(baruta_Site_indicators_ps_object_tax_B_merged$Class)]<-"Unknown"
baruta_Site_indicators_ps_object_tax_B_merged$Family[is.na(baruta_Site_indicators_ps_object_tax_B_merged$Family)]<-"Unknown"
baruta_Site_indicators_ps_object_B_merged@tax_table<-tax_table(baruta_Site_indicators_ps_object_tax_B_merged)
fam_names<-data.frame(gsub("column", "", tax_table(baruta_Site_indicators_ps_object_B_merged)[,3]))
#heatmap with taxa names
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B_merged, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = tax_table(baruta_Site_indicators_ps_object_B)[,4], row_dend_reorder =TRUE, col= col_fun)
#heatmap with OTUids
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B_merged, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = taxa_names(baruta_Site_indicators_ps_object_B), row_dend_reorder =TRUE, col= col_fun)
```

Merge by treatment site and rerun algo.  No significant indicators.  
```{r}
#subset to post-treatment samples only
bb16S_bac_ML_B<-subset_samples(bb16S_bac_ML, Timepoint == "B")
#summarize by Treatment_Site
bb16S_bac_ML_B_merged<-merge_samples(bb16S_bac_ML_B, "Treatment_Site")
sample_data(bb16S_bac_ML_B_merged)$Treatment_Site<-rownames(sample_data(bb16S_bac_ML_B_merged))
#run algo and filter phyloseq object
bb16S_bac_ML_B_merged<-prune_taxa(taxa_sums(bb16S_bac_ML_B_merged) > 0, bb16S_bac_ML_B_merged)
baruta_Site_indicators_B_merged<-baruta_phyloseq(physeq = bb16S_bac_ML_B_merged, GroupingVar = "Treatment_Site")
#no taxa indicative of treatment x site interaction.
#Stop this inquiry here. Do not run below becasue no taxa are identified. 
baruta_Site_indicators_ps_object_B_merged<-subset_taxa(bb16S_bac_ML_B_merged, taxa_names(bb16S_bac_ML_B_merged) %in% baruta_Site_indicators_B_merged)
#extract taxonomy and rename NA to unknown. 
baruta_Site_indicators_ps_object_tax_B_merged<-tax_table(baruta_Site_indicators_ps_object_B_merged)
baruta_Site_indicators_ps_object_tax_B_merged<-data.frame(baruta_Site_indicators_ps_object_tax_B_merged)
baruta_Site_indicators_ps_object_tax_B_merged$Class[is.na(baruta_Site_indicators_ps_object_tax_B_merged$Class)]<-"Unknown"
baruta_Site_indicators_ps_object_tax_B_merged$Family[is.na(baruta_Site_indicators_ps_object_tax_B_merged$Family)]<-"Unknown"
baruta_Site_indicators_ps_object_B_merged@tax_table<-tax_table(baruta_Site_indicators_ps_object_tax_B_merged)
fam_names<-data.frame(gsub("column", "", tax_table(baruta_Site_indicators_ps_object_B_merged)[,3]))
#heatmap with taxa names
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B_merged, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = tax_table(baruta_Site_indicators_ps_object_B)[,4], row_dend_reorder =TRUE, col= col_fun)
#heatmap with OTUids
micrUBIfuns::plot_taxa_heatmap(baruta_Site_indicators_ps_object_B_merged, rm_na = F,scale_by = "taxa", cluster_rows=F, cluster_columns = T ,column_labels = taxa_names(baruta_Site_indicators_ps_object_B), row_dend_reorder =TRUE, col= col_fun)
```


Exploratory tree of indicators from initial Baruta and then split by timpeoints. 
```{r}
plot_tree(baruta_Site_indicators_ps_object, color="Site", shape = "Timepoint", label.tips = "taxa_names")

plot_tree(baruta_Site_indicators_ps_object_A, color="Site", shape = "Treatment", label.tips = "taxa_names")

plot_tree(baruta_Site_indicators_ps_object_B, color="Site", shape = "Treatment", label.tips = "taxa_names")
```
Shared and unique taxa among sites
```{r}
unique_taxa(phyloseq_obj = bb16S_bac_rarefy, treatment = "Site")
unique_taxa(phyloseq_obj = bb16S_bac_rarefy, treatment = "Region")

common_taxa(phyloseq_obj = bb16S_bac_rarefy, treatment = "Site", n = 6)
common_taxa(phyloseq_obj = bb16S_bac_rarefy, treatment = "Site", n = 4)
```

